---
title: 'Lab 8: Clustering Analysis'
author: "Meghan Fletcher"
date: "2/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)

# Packages for cluster analysis:
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)

```

## Part 1: K-means Clustering

#### Exploratory Visualization
```{r}
# bill length vs. depth exploratory plot:
ggplot(penguins) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange", "cyan4", "darkmagenta"))

# Flipper length vs. body mass exploratory plot:
ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange", "cyan4", "darkmagenta"))
  
```

for k-means clustering you need to specify the number of clusters *a priori*. R **does** have some tools to help you decide, but this should NOT override your judgement based on conceptual or expert understanding. 

```{r}
# How many clusters do you THINK there should be? 
number_est <- NbClust(penguins[3:6], min.nc = 2, max.nc = 10, method = "kmeans")

number_est

# By these estimators, 2 is identified as the best number of clusters by the largest number of algorithms (8 / 30)...but should that change our mind? Maybe...but here I think it makes sense to still stick with 3 (a cluster for each species) and see how it does. 
```

#### Create a complete, scaled version of the data

```{r}
# Drop rows where any of the four size measurements are missing
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

# Only keep the columns for the four size measurements, then SCALE them
penguins_scale <- penguins_complete %>% 
  select(ends_with("mm"), body_mass_g) %>% 
  scale()
```

#### Run k-means

```{r}
penguins_km <- kmeans(penguins_scale, 3) # kmenas specifying 3 groups to start
```

```{r}
# See what it returns (different elements returned by kmeans function):
penguins_km$size # How many observations assigned to each cluster
penguins_km$cluster # What cluster each observation in penguins_scale is assigned to

# Bind the cluster number to the original data used for clustering, so that we can see what cluster each penguin is assigned to
penguins_cl <- data.frame(penguins_complete, cluster_no = factor(penguins_km$cluster))

# Plot flipper length versus body mass, indicating which cluster each penguin is assigned to (but also showing the actual species):
ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = cluster_no,
                 shape = species))

# plot bill dimensions and mapping species & cluster number to the point shape and color aesthetics:
ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = cluster_no,
                 shape = species))
```

```{r}
# Find the counts of each species assigned to each cluster, then pivot_wider() to make it a contingency table:
penguins_cl %>% 
  count(species, cluster_no) %>% 
  pivot_wider(names_from = cluster_no, values_from = n) %>% 
  rename('Cluster 1' = '1', 'Cluster 2' = '2', 'Cluster 3' = '3')
```

Takeaway: as we see from the graph, *most* chinstraps in Cluster 1, and *most* Adelies in Cluster 2, and *all* Gentoos are in Cluster 3 by k-means clustering. So this actually does a somewhat decent job of splitting up the three species into different clusters, with some overlap in Cluster 1 between Adelies & chinstraps, which is consistent with what we observed in exploratory data visualization. 

## Part 2: hierarchical Clustering

#### Read in the WroldBank data and simplify to keep only top 20 GHG emitters
```{r}
wb_env <- read_csv("wb_env.csv")

wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)
```

#### Scale the data
```{r}
# Scale the numeric variables (columns 3:7)
wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()

# Update to add rownames (country name) from wb_ghg_20
rownames(wb_scaled) <- wb_ghg_20$name
```

#### Find the Euclidian distances

```{r}
# Compute dissimilarity values (Euclidean distances):
euc_distance <- dist(wb_scaled, method = "euclidian")

#view(euc_distance)
```

#### Perform hierarchical clustering by complete linkage with `stats::hclust()`

```{r}
# Hierarchical clustering (complete linkage)
hc_complete <- hclust(euc_distance, method = "complete")

plot(hc_complete, cex = 0.6, hang = -1)
```

#### Now try it with sinle linkage & compare

```{r}
# Hierarchical clustering (single linkage)
hc_single <- hclust(euc_distance, method = "single")

plot(hc_single, cex = 0.6, hang = -1)
```

#### Make a tanglegram to compare dendrograms 

Use the `dendextend::tanglegram()` function to make it
First, convert to class `dendrogram`, then combine them into a list:

```{r}
# Convert to class dendrogram
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)
```

```{r}
# Make a tanglegram
tanglegram(dend_complete, dend_single)
```

#### Plot dendrogram with ggplot

```{r}
ggdendrogram(hc_complete,
             rotate = TRUE) +
  theme_minimal() +
  labs(x = "country")
```


